---
title: "Basic overview of subset of OBIS occurrence data"
author: ~
date: '2019-07-23'
categories: []
tags: []
---

```{r mpa compare imports, include=FALSE}
library(glue)
# also uses robis
source("./../R/obis_cache.R")
```


![Plots of reef fish indices inside (protected) and outside (not protected) no take marine zones across years in the Florida Keys. The units for panel A are abundance (number of individuals), panel B is kilograms per 40,000 km2, panel C is units ranging from 0 (maximally uneven) to 1 (maximally even), and panels D-G are in units of effective number of species. From Hepner, M. E. (2017). Reef Fish Biodiversity in the Florida Keys National Marine Sanctuary.](mhepner_fig6.png)

```{r mpa areas}
fknms <- "(-83.15 25.65, -80.066667 25.65, -80.066667 24.3, -83.15 24.3, -83.15 25.65)"
# special use areas manually transcribed from images at
# https://floridakeys.noaa.gov/zones/special/welcome.html
# converted from min/degrees to decimal degrees using
# https://www.pgc.umn.edu/apps/convert/
# https://floridakeys.noaa.gov/zones/special/easternsambo.html
eastern_sambo <- "(-81.667167 24.496167, -81.659833 24.497333, -81.655833 24.4925, -81.666 24.4895, -81.667167 24.496167)"
# TODO: include others
# https://floridakeys.noaa.gov/zones/emas/looekey.html
# https://floridakeys.noaa.gov/zones/special/tennessee.html
# https://floridakeys.noaa.gov/zones/special/conchreef.html
# TODO add others here too
inside_mpas <- glue("MULTIPOLYGON (({eastern_sambo}))")
outside_mpas <- glue("MULTIPOLYGON (({fknms},{eastern_sambo}))")
```

```{r fetch mpa compare data}
species_name <- "Ocyurus chrysurus"
species_name_safe <- sanitize_query_id(species_name)
inside_query_id  <- glue( "mpa_inside_{species_name_safe}")
outside_query_id <- glue("mpa_outside_{species_name_safe}")
if (has_cache(inside_query_id)){
    inside_occurrence <- load_cache(inside_query_id)
} else {
    inside_occurrence <- robis::occurrence(
        species_name,
        geometry = inside_mpas
    )
    save_cache(inside_query_id, inside_occurrence)
}
if (has_cache(outside_query_id)){
    outside_occurrence <- load_cache(outside_query_id)
} else {
    outside_occurrence <- robis::occurrence(
        species_name,
        geometry = outside_mpas
    )
    save_cache(outside_query_id, outside_occurrence)
}
```

```{r mpa compare timeseries, echo=FALSE}
# TODO: how to make this one plot?
ggplot2::ggplot(inside_occurrence) + ggplot2::geom_bar(ggplot2::aes(date_year), width = 1)
ggplot2::ggplot(outside_occurrence) + ggplot2::geom_bar(ggplot2::aes(date_year), width = 1)
```
